---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
library(simsurv)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# simsurv

[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/simsurv)](http://www.r-pkg.org/pkg/simsurv) 
[![License](https://img.shields.io/badge/License-GPL%20%28%3E=%203%29-brightgreen.svg)](http://www.gnu.org/licenses/gpl-3.0.html)

**simsurv** is an R package that allows the user to simulate survival times from
standard parametric survival distributions (exponential, Weibull, and Gompertz),
2-component mixture distributions, or a user-defined hazard function, log hazard
function, cumulative hazard function, or log cumulative hazard function.
Baseline covariates can be included under a proportional hazards assumption.
Time dependent effects (i.e. non-proportional hazards) can be included by 
interacting covariates with linear time or some transformation of time.

Under the 2-component mixture distributions the baseline survival at time 
*t* is taken to be *S(t) = p x S_1(t) + (1 - p) x S_2(t)*
where *S_1(t)* and *S_2(t)* are the baseline survival under each component of 
the mixture distribution and *p* is the mixing parameter. Each component of the 
mixture distribution is assumed to be either exponential, Weibull or Gompertz.
The 2-component mixture distributions can allow for a variety of flexible
baseline hazard functions (see Crowther and Lambert (2013) for some examples).

If the user wishes to provide a user-defined [log] [cumulative] hazard function 
(instead of using one of the standard parametric survival distributions) then 
this is also possible. 
If a user-defined hazard or log hazard function is specified, then this 
is allowed to be time-dependent, and the resulting cumulative hazard function
does not need to have a closed-form solution. The survival times are
generated using the approach described in Crowther and Lambert (2013),
whereby the cumulative hazard is evaluated using numerical quadrature and
survival times are generated using an iterative algorithm which nests the
quadrature-based evaluation of the cumulative hazard inside Brent's (1973)
univariate root finder. Not requiring a closed form solution to the cumulative
hazard function has the benefit that survival times can be generated for
complex models such as flexible (spline-based) baseline hazards or under
joint longitudinal and survival models; the package documentation and vignettes
provide examples of this.

For further details on the underlying methodology and examples of usage, please
see the package vignettes (available 
[here](https://cran.r-project.org/web/packages/simsurv/index.html)).

Note that this package is modelled on the user-written **survsim** package
available in the Stata software (see Crowther and Lambert (2012)).

**Note:** Please note that the version available on GitHub is the most up-to-date 
*development* version of the package. A stable version of the package is
available from the Comprehensive R Archive Network (CRAN); type 
`install.packages("simsurv")` into your R session to download it.

## Getting Started

### Installation: stable release

You can install the **simsurv** package directly from CRAN. Just type the
following into your R session console:

```r
install.packages("simsurv")
```
### Installation: development version

If, for some reason, you wish to download the development version from GitHub,
then this can be done easily using the **devtools** package. 
To do this you should first check you have devtools installed by executing the 
following commands from within your R session:

```r
if (!require(devtools)) {
  install.packages("devtools")
}
```

Then execute the following commands to install **simsurv**:

```r
library(devtools)
install_github("sambrilleman/simsurv")
```

## Examples

Please also see the package [vignettes](https://cran.r-project.org/web/packages/simsurv/index.html) for more detailed examples and a
description of the methodology underpinning the **simsurv** package.

### Simpler examples

Generate times from a Weibull model including a binary treatment variable, with log hazard ratio of -0.5, and censoring after 5 years:

```{r example1}  
set.seed(13579)
covs <- data.frame(id = 1:1000, trt = stats::rbinom(1000, 1L, 0.5))
s1 <- simsurv(lambdas = 0.1, gammas = 1.5,
              x = covs, betas = c(trt = -0.5), maxt = 5)
head(s1)
```

Generate times from a Gompertz model:

```{r example2}
s2 <- simsurv(dist = "gompertz", lambdas = 0.1, gammas = 0.05, x = covs)
```

Generate times from a 2-component mixture Weibull model:

```{r example3}
s3 <- simsurv(lambdas = c(0.1, 0.05), gammas = c(1, 1.5),
              mixture = TRUE, pmix = 0.5, x = covs, maxt = 5)
```

Generate times from user-defined log hazard function:

```{r example4}
fn <- function(t, x, betas, ...)
  (-1 + 0.02 * t - 0.03 * t ^ 2 + 0.005 * t ^ 3)
s4 <- simsurv(loghazard = fn, x = covs, maxt = 1.5)
```

### Complex examples

In this section we present an example showing how the **simsurv** package's main function can be used to simulate survival times based on a joint longitudinal and survival model. (In most instances, simulating survival times from a joint longitudinal and survival model can pose difficulties since time-dependency in the hazard function leads to there not being a general closed form solution to the integral when evaluating the cumulative hazard). 

First we define the hazard function to pass to `simsurv`. This function must be defined by the user and must use three named arguments: `t`, which is the time variable; `x`, which is a data frame with each row containing the covariate values for one individual; `betas`, which is a data frame with each row containing the "true" parameter values for one individual. The function must return the hazard evaluated at time `t`. We will define the hazard function for a Weibull proportional hazards regression submodel from a joint longitudinal and survival model with a so-called "current value" association structure, where the hazard at time *t* depends on the expected value of the longitudinal outcome also evaluated at time *t*.

```{r hazfn, eval=T}
haz <- function(t, x, betas, ...) {
  betas[["shape"]] * (t ^ (betas[["shape"]] - 1)) * exp(
    betas[["betaEvent_intercept"]] +
    betas[["betaEvent_binary"]] * x[["Z1"]] +
    betas[["betaEvent_continuous"]] * x[["Z2"]] +
    betas[["betaEvent_assoc"]] * (
      betas[["betaLong_intercept"]] +
      betas[["betaLong_slope"]] * t +
      betas[["betaLong_binary"]] * x[["Z1"]] +
      betas[["betaLong_continuous"]] * x[["Z2"]]
    )
  )
}
```

Then we construct data frames with the true parameter values and the covariate data for each individual:

```{r parscovs, eval=T}
set.seed(5454) # set seed before simulating data
N <- 20        # number of individuals

# Population (fixed effect) parameters
betas <- data.frame(
  shape                = rep(2,    N),
  betaEvent_intercept  = rep(-11.9,N),
  betaEvent_binary     = rep(0.6,  N),
  betaEvent_continuous = rep(0.08, N),
  betaEvent_assoc      = rep(0.03, N),
  betaLong_binary      = rep(-1.5, N),
  betaLong_continuous  = rep(1,    N),
  betaLong_intercept   = rep(90,   N),
  betaLong_slope       = rep(2.5,  N)
)

# Individual-specific (random effect) parameters
b_corrmat <- matrix(c(1, 0.5, 0.5, 1), 2, 2)
b_sds     <- c(20, 3)
b_means   <- rep(0, 2)
b_z       <- MASS::mvrnorm(n = N, mu = b_means, Sigma = b_corrmat)
b         <- sapply(1:length(b_sds), function(x) b_sds[x] * b_z[,x])
betas$betaLong_intercept <- betas$betaLong_intercept + b[,1]
betas$betaLong_slope     <- betas$betaLong_slope     + b[,2]

# Covariate data
covs <- data.frame(
  Z1 = rbinom(N, 1, 0.45), # a binary covariate
  Z2 = rnorm(N, 44, 8.5)   # a continuous covariate
)
```

Then we simulate the survival times based on the hazard function (passed to the `hazfn` argument), the covariate values for each individual (passed to the `x` argument), and true parameter values for each individual (passed to the `betas` argument). We will also censor individuals with survival times greater than 10 years (or whatever time units we have assumed to be using) by specifying the `maxt` argument:

```{r simsurv, eval=T}
s1 <- simsurv(hazard = haz, x = covs, betas = betas, maxt = 10)
head(s1)
```

## Bug Reports

If you find any bugs, please [file an issue on GitHub](https://github.com/sambrilleman/simsurv/issues) or report them via email to [Sam Brilleman](mailto:sam.brilleman@monash.edu).

## References

1. Crowther MJ, and Lambert PC. Simulating biologically plausible complex survival data. *Statistics in Medicine* 2013; **32**, 4118--4134. doi:10.1002/sim.5823

2. Bender R, Augustin T, and Blettner M. Generating survival times to simulate Cox proportional hazards models. *Statistics in Medicine* 2005; **24(11)**, 1713--1723.

3. Brent R. (1973) *Algorithms for Minimization without Derivatives*. Englewood Cliffs, NJ: Prentice-Hall.

4. Crowther MJ, and Lambert PC. Simulating complex survival data. *The Stata Journal* 2012; **12**(4), 674--687. http://www.stata-journal.com/sjpdf.html?articlenum=st0275
